GRN inference in multiple species

### 2015-04-06 paper1.tex

\subsection{Fused GRN}
We expect related species to be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_S(\beta^A_{g,k} - \beta^B_{h,l})^2$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} \lambda_S(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})^2
\end{equation}
Where the second sum is over pairs of interactions with fusion constraints. Note that each term in the quantity being minimized is an L2 norm of a linear combination of gene expressions, so the entire problem can be solved through linear regression with a suitable design matrix [DETAILS SHOULD GO SOMEWHERE ELSE, OR NOWHERE]. The $\beta$ vectors corresponding to a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

@@

\subsection{Alternative penalty functions}
Sometimes gene orthology is not a good predictor of functional similarity. To simultaneously learn when a gene's function has diverged from that of its ortholog, and improve model selection, we consider alternative penalized least squares approaches. Our fusion penalty pulls beta values for analogous interactions towards each other; since we use an L2 penalty, interaction weights which differ from each other the most are excessively penalized. SCAD [Fan Li 2001] and MCP [Zhang] are two alternative penalties which diminish this bias. Fan and Li point out that the bias can be removed by choosing a constant penalty beyond some threshold. This is analogous to 'unfusing' certain interactions, and easing up on the demand that certain beta values be close in value when doing so strongly opposes minimizing the mean squared error component of the penalty function. SCAD and MCP both use an L1 penalty near the origin, setting small estimated coefficients to zero and choosing a sparse solution. Because we do not expect most fused beta values to be identical, but rather similar, we amend the alternative fusion penalty to be L2 near the origin, resulting in:
\begin{equation}
p_{\lambda,\gamma}(\theta) = \left\{
    \begin{array}{lr}
    \lambda\theta^2-{ \theta^2 \over 2\gamma} & \text{if } \theta \leq \gamma\lambda\\
    {1 \over 2}\gamma\lambda^2 & \text{if } \theta > \gamma\lambda
    \end{array}
    \right. 
\end{equation}
It is more intuitive to think about the derivative of the penalty:
\begin{equation}
p'_{\lambda,\gamma}(\theta) = \left\{
    \begin{array}{lr}
    2\lambda\theta - {\theta \over \gamma} & \text{if } \theta \leq \gamma\lambda\\
    0 & \text{if } \theta > \gamma\lambda
    \end{array}
    \right.
\end{equation}

The MCP-like penalty begins by applying the same rate of penalization as the L2, but continuously relaxes the penalization until the rate drops to 0 and the penalty saturates. 
As in [Fan Li], we use local quadratic approximation and solve by iteratively computing ridge regression.

@@

\subsection{Bacterial data}
We

is good
Look an equation array
\begin{equation}
\begin{array}{l}
R_i^{\mathrm{input}}(\theta) = c g_i^{\mathrm{input}}\exp(-(\theta - \theta_i)^2 / (2\sigma^2_{\mathrm{input}}))
\\
R_i^{\mathrm{output}}(\theta) = g_i^{\mathrm{output}}\sum_{j=0}^N R_j^{\mathrm{input}}(\theta) \exp(-(\theta_i - \theta_j)^2/(2\sigma^2_{\mathrm{output}})) .
\end{array}
\end{equation}
\bibliographystyle{plain}
\bibliography{paper}

>>>>

\subsection{Fused GRN}
Related species should be governed by similar but not necessarily identical gene regulatory networks. We approach the problem of cross-species GRN inference by introducing constraints into the above regression formulation to penalize differences between interaction weights that are expected to be similar based on gene orthology. We can then solve the penalized regression problems simultaneously, in order to obtain a GRN for each species. Consider the case of organisms $A$ and $B$, governed by GRNs $\beta^A$ and $\beta^B$ (the following approach applies equally well to more than two species but for simplicity we continue with the case of two species). If TF $g^A$ in organism $A$ and TF $h^B$ in organism $B$ are orthologs, and gene $k^A$ and $l^B$ are orthologs, then we expect that the $g^A \rightarrow k^A$ interaction weight should be similar to the $h^B \rightarrow l^B$ interaction weight, and we introduce a fusion constraint between these analogous interactions. In terms of the above regression formulation, we expect that $\beta^A_{g,k} \approx \beta^B_{h,l}$, and include a penalty term $\lambda_S(\beta^A_{g,k} - \beta^B_{h,l})^2$ to the quantity being minimized in order to ensure similarity. The parameter $\lambda_S$ controls how highly to penalize differences between fused coefficients. This gives us the final equation to be minimized 
\begin{equation}
\argmin_{(\beta^A, \beta^B)} \displaystyle\sum_{S \in (A, B)} \vert \vert X^S\beta^S - Y^S \vert \vert ^2 + \lambda_R \vert \vert \beta^S \vert \vert ^2 + \displaystyle \sum_{\beta^{S_1}_{g,k} \approx \beta^{S_2}_{h,l}} \lambda_S(\beta^{S_1}_{g,k} - \beta^{S_2}_{h,l})^2
\end{equation}

Where the second sum is over pairs of interactions with fusion constraints. Note that each term in the quantity being minimized is an L2 norm of a linear combination of gene expressions, so the entire problem can be solved through linear regression with a suitable design matrix [DETAILS SHOULD GO SOMEWHERE ELSE, OR NOWHERE]. The regulators of a pair of genes can be solved independently as long as the pair is not linked by fusion constraints, or a chain of fusion constraints; thus the problem of cross-species GRN inference decomposes into a large number of manageable regression problems as long as the orthology mappings are sparse.

@@

\subsection{Alternative penalty functions}
Fusion constraints are intended to penalize dissimilarity between interactions thought to be evolutionarily analogous. Because fusion constraints are L2, interaction weights which differ from each other by a large amount are excessively penalized, which effectively ensures that fused interactions are assigned similar weights. This may be inappropriate for interactions which are identified based on orthology as being analogous, but which are no longer similar due to evolutionary changes. We propose that a saturating penalty may be useful for dealing with uncertainty about which interactions are conserved.

With a saturating fusion penalty, fused interactions which appear to be very different based on expression data are allowed to unfuse from one another. A closely related problem has been studied in the context of LASSO regularization, where it was shown by Fan and Li that using a saturating penalty retains many of LASSO's desireable properties while removing its bias towards 0. They further showed that, although the resulting loss-function is nonconvex, good results can be obtained with a local quadratic approximation of gradient descent. Several saturating penalties, such as SCAD [Fan Li 2001] and MCP [Zhang], have been discussed in the context of sparse regression. We introduce a modified form of MCP to the problem of penalizing differences between fused coefficients. The principle difference between the penalty we adopt and SCAD/MCP is that both of these penalties are L1 like at the origin, producing sparse solutions. Because we are penalizing differences in interaction weights, rather than the weights themselves, there's no particular reason to prefer sparsity.

We use a penalty on the difference between fused coefficients $\theta$ which is L2 like at the origin, begins to level out at $\theta = \frac{a}{2}$, and saturates at $\theta = a$. Written in terms of its derivative, the penalty $p'_{\lambda, a}$

\begin{equation}
p'_{\lambda,a}(\theta) = \left\{
    \begin{array}{lr}
    2\lambda\theta & \text{if } \theta \leq \lambda\\
    \text{max}(\lambda_S(a-\theta),0) & \text{if } \theta > {a \over 2}
    \end{array}
    \right.
\end{equation}

We also implement MCP [IS THIS MODIFIED TO BE L2 like at the origin?]

\begin{equation}
p_{\lambda,\gamma}(\theta) = \left\{
    \begin{array}{lr}
    \lambda\theta^2-{ \theta^2 \over 2\gamma} & \text{if } \theta \leq \gamma\lambda\\
    {1 \over 2}\gamma\lambda^2 & \text{if } \theta > \gamma\lambda
    \end{array}
    \right. 
\end{equation}

With derivative

\begin{equation}
p'_{\lambda,\gamma}(\theta) = \left\{
    \begin{array}{lr}
    2\lambda\theta - {\theta \over \gamma} & \text{if } \theta \leq \gamma\lambda\\
    0 & \text{if } \theta > \gamma\lambda
    \end{array}
    \right.
\end{equation}

As in [Fan Li], we use solve using iterative local quadratic approximation. Specifically,

\begin{equation}
\begin{array}{l}
\beta^S(t)\text{ is the network on iteration } t
\\
\text{for each fused  } B^{S_1}_{g,k} \approx B^{S_2}_{h,l}  \text{ we define } 
\\
\theta(0)=0
\\
\theta(t) = \vert B^{S_1}_{g,k} - B^{S_2}_{h,l} \vert
\\
\text{and introduce a fusion constraint } \lambda = \frac{p'(\theta(t))}{2\theta(t)} 
\\
\beta^S(t+1) \text{ is obtained by fitting the ridge-fused model with fusion constraints given by the above }\lambda s
\end{array}
\end{equation}

@@

\subsection{Bacterial data}
TBA

\bibliographystyle{plain}
\bibliography{paper}
